\textbf{ Matching the forecast horizon with the relevant spatial and temporal processes and data sources }

### 2019-12-09 doc/manuscript.tex
started new Intro paragr

\title{\textbf{ Matching the forecast horizon with the relevant ecological processes }}

@@

\noindent \emph{Data accessibility statement:} The manuscript contains no original data. All computer code is available at https://github.com/pbadler/space-time-forecast. Upon acceptance of the manuscript, code will be archived at Zenodo.

@@

\noindent \emph{Word counts:} 174 words in abstract; 4044 words in main text.

\noindent \emph{References}: 35

\noindent \emph{References}: 6 figures, no tables or text boxes

@@

@@

Most models used to generate ecological forecasts take either a time-series approach, based on long-term data from one location, or a space-for-time approach, based on data describing spatial patterns across environmental gradients. Here we consider how the forecast horizon determines whether the most accurate predictions come from the time-series approach, the space-for-time approach, or a combination of the two. We use two simulation case studies to show that forecasts for short and long-time scales need to focus on different ecological processes, which are reflected in different kinds of data. In the short-term, dynamics reflect initial conditions and fast processes such as birth and death, and the phenomenological time-series approach makes the best predictions. In the long-term, dynamics reflect the additional influence of slower processes such as evolutionary and ecological selection, colonization and extinction, which the space-for-time approach can effectively capture. At intermediate time-scales, a weighted average of the two approaches shows promise. However, making this weighted model operational will require new research to predict the rate at which slow processes begin to influence dynamics.

@@

Models used for ecological forecasting typically rely on either time-series approaches or space-for-time substitutions. The time-series approach involves fitting models to long-term datasets to describe the temporal dynamics of a system.  We then use those dynamic models to make predictions about what will happen in the future. This approach is often used to study population or vital rate fluctuations as a function of weather \citep{dalgleish_climate_2011}, or primary production as a function of annual precipitation \citep{lauenroth_long-term_1992}. Whether process-based or data-driven (e.g., \citealt{ward_complexity_2014}), time-series models capture ``fast processes'' operating on interannual time-scales, such as birth, death, individual growth, small-scale dispersal events, and short-term responses to environmental conditions (Fig. \ref{fig:fast-slow-concept}). However, models built using this approach normally cover a limited spatial extent (but see \citealt{Hefley2017,kleinhesselink_response_2018}), and ignore slower processes, such as evolutionary adaptation or turnover in community composition, that could influence dynamics at longer time scales \citep{clark_ecological_2001}.

@@

Whether historical dynamics, contemporary spatial patterns, or some combination of the two will serve as the best source of 
information for forecasting may depend on how far into the future we are attempting to forecast \citep{harris_forecasting_2018}. 
This potential dependency on the ``forecast horizon'' (\textit{sensu} \citealt{Hyndman2018})
reflects lags in the response of ecological conditions to environmental change, shifts in the importance of ecological processes with
time scale \citep{levin_1992,rosenzweig_1995}, and differences between time-series and spatial gradients in the range of 
environmental conditions represented in observed data (Fig. \ref{fig:fast-slow-concept}).
At short forecast horizons (days to years), dynamics will reflect the physiological and demographic responses of 
the organisms present at a site more than turnover of genotypes or species, 
environmental conditions are likely to stay within the range of historical variation, 
and the current state of the system is likely to capture the influence of unmeasured processes. As a result, 
for near-term forecasts time-series approaches may capture the key dynamics and provide accurate predictions.

In contrast, at long forecast horizons (decades to centuries),  environmental conditions that have not been 
historically observed are likely to not only occur but to persist long enough to drive significant turnover of genotypes and species 
along with changes in the flux of energy and nutrients.  At these long scales, the current state of the system may be 
little help in predicting the future state. For the century-scale forecasts often featured in biodiversity and 
species-distribution modeling, space-for-time approaches may effectively capture the response of ecosystems to major shifts 
in climate over long periods, producing better long-term forecasts than time-series approaches. 
Using different modeling approaches for different forecast horizons is common in other disciplines.
For example, meteorological models for short-term weather forecasts differ substantially in spatial
and temporal resolution and extent from the global circulation models used to predict long-term changes
in climate.

Here we use simulation models to 1) demonstrate that the best model-building approaches for ecological forecasting 
depend on the time horizon of the forecast, and 2) explore how time-series and space-for-time approaches might be 
combined via weighted averaging to make better forecasts at intermediate time scales. We conduct two simulation case studies, one 
focused on how interspecific interactions affect the population dynamics of a focal species, and the second focused on
an eco-evolutionary scenario. Our analyses show that: 
\begin{enumerate}
	\item For short-term forecasts, phenomenological time-series approaches are hard to beat, whereas longer-term forecasts require accounting for the influence of slow processes such as evolutionary and ecological selection as well as dispersal.
	\item Different kinds of data reflect the operation of different processes: longitudinal data capture autocorrelation and fast responses of current assemblages to interannual environmental variation, while data spanning spatial gradients capture the long-term outcome of interactions between fast and slow processes.  Whether predictive models should be trained using longitudinal or spatial data sets, or both, depends on the time-scale of the desired forecast.
	\item A key challenge for future research is determining the rate at which slow processes begin to influence dynamics.
\end{enumerate}

@@

In two case studies, we simulated the effects of an increase in temperature on simple systems with known dynamics.
The ``truth'' is represented by a model that is mechanistic for at least one important process, but we treat the model as unknown when
analyzing the data and we assumed that perfectly recovering this model would not be possible in practice. We began each simulation
under stationary temperature, allowing the system to equilibrate; we call this the baseline phase. 
We then increased temperature progressively over a period of time, followed by a second period of stationary, now elevated, temperature. 
The objective was to forecast the response of the system to the temperature increase based on data gathered during the baseline period.

We made forecasts based on two phenomenological models, each representing processes operating at different time scales. One model represents the time-series or ``temporal approach.'' We correlated interannual variation in an ecological response with interannual variation in the environment at just one site. The other model represents the space-for-time substitution approach, which we call the ``spatial approach'' for brevity. We correlated the mean environment with the mean of an ecological state or rate across many sites.  We compared forecasts from both models to the simulated dynamics to determine how well the two approaches performed at different forecast horizons. We also assessed the potential for combining the information available in temporal and spatial patterns by using a weighted average of the forecasts from the temporal and spatial approaches optimized to best match the (simulated) observations. We then studied how the optimal model weights changed over time. We expected the temporal model to best predict short-term dynamics, the spatial model to best predict long-term dynamics, while the weighted model would show potential to provide the best forecasts at transitional, intermediate time scales. The three statistical models are described in Supporting Information (Appendix \ref{models}) and all code for both case studies is available at Github (https://github.com/pbadler/space-time-forecast).

@@

Conservation biologists and natural resource managers often need to anticipate the impact of environmental change on the abundance of 
endangered species, biological invaders, and harvested species. Although the managers may be primarily interested in just one focal species, skillful prediction might require considering interactions with many other species, greatly complicating the problem. But at what forecast horizon do altered species interactions become impossible to ignore? We explored this question using a metacommunity model developed by \cite{alexander_lags_2018} to study how community responses to increasing temperature depend on the interplay between within-site demography and competitive interactions and the movement of species across sites. The model features Lotka-Volterra competitive interactions among plants within sites that are arrayed along an elevation and temperature gradient. Composition varies along the gradient because of a trade-off between growth rate and cold tolerance: cold sites are dominated by slow-growing species that can tolerate low temperatures, while warm sites are dominated by  fast-growing species that are cold intolerant. Multiple species can coexist within sites because all species experience stronger competition from conspecifics than from heterospecifics. Sites are linked by dispersal: a specified fraction of each species' offspring leaves the site where they were produced and reaches all other sites with equal probability. We provide a more detailed description of the model in SI Appendix \ref{metacomm}.

@@

During the baseline period there were strong spatial patterns across the mean temperature gradient. 
Individual species, including our focal species, showed classic, unimodal ``Whittaker" patterns of abundances across the gradient (Fig. \ref{fig:species-patterns-models}A).
These spatial patterns are the basis for our ``spatial model" of the temperature-biomass relationship for our focal species (Fig. \ref{fig:species-patterns-models}A). 
In contrast to the strong spatial patterns, population and community responses to interannual variation in temperature within sites were weak.
At our focal site in the center of the gradient, the biomass of the focal species was quite insensitive to interannnual variation in temperature, but showed strong temporal autocorrelation (Fig. \ref{fig:species-patterns-models}B). Our ``temporal model" estimates this weak, linear temperature effect, along with the strong lag effect of biomass in the previous year.

We fit both a temporal and a spatial statistical model to forecast the effect of a temperature increase (Fig. \ref{fig:community-forecast-species}A) on the focal species' biomass at one location in the center of the temperature gradient. 
The predictions from the spatial and temporal models contrasted markedly, with the temporal 
model predicting a large increase in biomass and the spatial model predicting a decrease. Initially, the simulated abundances followed the increase predicted by the 
temporal model, but as faster-growing species colonized and increased in abundance at the focal site, the biomass of the focal species decreased, eventually falling below its baseline level (Fig. \ref{fig:community-forecast-species}B).

To combine the temporal and spatial model into a single forecast, we fit a weighting parameter, $\omega$, which varies over time and is bounded between 0 and 1. At any time point, $t$, this weighted forecast is $\omega \cdot T(N_{t-1},K_t) + (1-\omega) \cdot S(K_t) $ where $T$ is the temporal model, which depends on population size, $N$, and expected temperature, $K$, and $S$ is the spatial model, which depends only on $K$ (see SI Appendix \ref{models} for a full description of the approach). The weighted model accurately predicts the simulated dynamics across the full forecast horizon (Fig. \ref{fig:community-forecast-species}B). It also shows that the most rapid shifts in the model weights occurred during the period when warm-adapted, faster growing species were increasing most rapidly in abundance (Fig. \ref{fig:community-forecast-species}C). However, the reason the weighted models works so well is that the weights were determined by fitting directly to the data. Unlike our spatial and temporal model forecasts, we did not generate out-of-sample predictions from the weighted model; it merely provides a convenient way to quantify how rapidly dynamics shift from being dominated by interannual variation captured in the temporal model (time $t=0$ to $t \approx 1250$ in Fig. \ref{fig:community-forecast-species}B) to being dominated by the steady-state equilibrium captured by the spatial model (time $t \geq 2500$). A true forecast from the weighted model would require a method to determine the model weights \emph{a priori}.

The compositional turnover affecting our focal species also influences total biomass, linking community and ecosystem dynamics. We repeated our focal species analysis for total community biomass, and the results were similar: the temporal model initially made the best forecasts immediately following the onset of the temperature increase, but as the identity and abundances of species at the study site changed, the model weights rapidly shifted to the spatial model (SI Figs. \ref{fig:community-models-total} and \ref{fig:community-forecasts-total}).

@@

Evolutionary adaptation is a key uncertainty in predicting how environmental change will impact a focal population at a given location \citep{Hoffmann2011}. Like the shifts in species composition illustrated in the previous example, shifts in genotype frequencies can also influence dynamics and forecasts at different time scales. Although shifts in genotype frequencies at the population level are analogous to changes in species composition at the community level, the mechanisms are distinct: heterozygosity and genetic recombination have no analogue at the community level. We demonstrate how these processes influence short and long-term forecasts with a standard eco-evolutionary model for a hypothetical annual plant population in which fecundity is temperature dependent, and different genotypes have different temperature optima (Fig. \ref{fig:evo_evo_spatial_temporal_models}A). Our model describes how the local density of each genotype changes between years,  which depends on temperature and genotype densities in the previous year. Transient temporal dynamics are computed directly from the model; these dynamics define our temporal forecast. To create ``spatial data'', we simulated the equilibrium density of each genotype under different mean temperatures. The pattern of equilibrium densities across a gradient in mean annual temperature defines our spatial forecast: cold sites will be dominated by the cold-adapted homozygous genotype, warm sites will be dominated by the heat-adapted homozygous genotype, and intermediate sites will be dominated by the heterozygous genotype (Fig. \ref{fig:evo_evo_spatial_temporal_models}B). The full model description is provided in SI Appendix \ref{eco-evo}.

@@

To demonstrate these dynamics, we simulated a diploid annual plant population at a colder than average site. During the baseline period, the population is dominated by the cold-adapted genotype. We used the simulated data from this baseline period to fit an empirical model that assumes no knowledge of the underlying eco-evolutionary process. This empirical temporal model (Appendix \ref{models}) predicts population growth rate as a function of annual temperature and population size (Fig. \ref{fig:evo_evo_spatial_temporal_models}C). We then imposed a period of warming, followed by a final period of higher stationary temperature (Fig. \ref{fig:forecast} top).

With the onset of warming, the population crashed as the cold-adapted genotype decreased in abundance. Eventually, frequencies of the heterozygous genotype and the warm-adapted homozygous genotype began to increase and the population recovered (Fig. \ref{fig:forecast} bottom). The temporal model (solid blue line in Fig. \ref{fig:forecast}) accurately predicted the impact of the initial warming trend, but eventually became too pessimistic, while the spatial model (solid red line in Fig. \ref{fig:forecast}) did not handle the initial trend but accurately predicted the eventual, new steady state (Fig. \ref{fig:forecast} bottom).

As in the community turnover example, we also fit a weighted average of the spatial and temporal model, with the weights changing over time. This weighted model initially reflected the temporal model (decrease from $t=500$ to $t=600$), but then rapidly transitioned to reflect the spatial model ($t \geq 700$). The rapid transition in the weighting term, $\omega$, occurred during the period of most rapid change in genotype frequencies (Fig. \ref{fig:forecast_supp}). The weighted model's predictions look impressively accurate, but, as in the community turnover example, that is because we used the full, simulated time series to fit the weighting term. A true forecast would require an independent method to predict how the model weights shift over time.

@@

Ecological forecasts are typically made using either a space-for-time substitution approach based on models fit to spatial data or using dynamic models fit to time-series data. Our results demonstrate that these two approaches can make very different predictions about the future state of ecological systems. Which approach provides the most accurate forecasts depends on the forecast-horizon. In our simulations, time-series approaches performed best for short-term forecasts, whereas models based on spatial data made more accurate long-term forecasts. In addition, our simulations demonstrate extended transitional periods during which neither the time-series or the spatial approach is effective on its own. The challenge is determining what is ``short-term,'' what is ``long-term,'' and how to handle the many forecasts we need in ecology which fall in between. We have proposed that a weighted combination of the time-series and space-for-time approaches may produce better forecasts at these intermediate forecast horizons.

We designed our simulation studies to illustrate how the change in model performance with increasing forecast horizon reflects differences in the types and scales of processes captured by spatial and temporal data sets. How could these hypotheses be tested with empirical data? The hypothesis that time-series models will be most effective for near-term forecasts already has empirical support, in the form of recent analyses of biodiversity forecasts at time scales from one to ten years \citep{harris_forecasting_2018}. The result should not be surprising, since local time-series data capture demographic processes, lagged effects, and responses of current assemblages to small changes in environmental conditions. In addition, the state of the system in the near future depends heavily on the current state. Since short-term forecasts do not typically require extrapolating into novel conditions, a model based on the historical range of variation which incorporates lags and accurate initial conditions is likely to be successful.

@@

The greatest empirical challenge will be testing our hypothesis that a weighted average of spatial and temporal models will make the best forecasts at intermediate time scales. There are two problems: finding appropriate data and determining the model weights \emph{a priori}. Many data sets have both a longitudinal and spatial dimension, but we could not think of one which also featured a clear ecological response to significant environmental change. Surely such datasets exist, and we hope researchers who work with them will test our proposed weighted model. Determining model weights may be more difficult. In our simulations, we fit the weights directly to the simulated data, which is impossible to do for actual forecasting when the future is unknown. We need new theory or empirical case studies in order to assign these weights \emph{a priori}.

@@

An alternative to a weighted combination of predictions from time-series and space-for-time models is to rely on fully process-based models. If we could accurately characterize all of the processes governing a system, then a model based on that understanding should make accurate predictions at all time-horizons. For example, rather than fitting phenomenological models to our simulated time-series, we could have fit the exact process-based models that we used to generate those time-series. Assuming reasonable estimates of the parameter, those models would have accurately predicted the dynamics at all forecast horizons. Process-based models should also be more robust for making predictions outside of historically observed conditions and even beyond the conditions observed across spatial gradients, which will be especially important for making predictions in a future with increasingly novel combinations of environment and species interactions \citep{Williams2007}. Unfortunately, in most cases this approach is not currently feasible because we lack a detailed knowledge of all the complex and interacting processes influencing the dynamics of real ecological systems. Even if the general form of the models was known, estimating the high number of parameters and quantifying how they vary across ecosystems typically requires more data than is currently available even for well studied systems. 
% mention difficulty of dealing with error propogation?
As a result, models used for ecological forecasting will include at least some phenomenological components. But that does not mean that phenomenological forecast models do not benefit from process-based understanding. The message from our simulations is that different processes should be considered for different forecast time-scales, and this can be done by fitting models to different kinds of datasets. Even when process-level understanding does not enable a fully mechanistic model, it can improve the specification of phenomenological models.

@@

Our results have important implications for the emerging field of ecological forecasting. First, they suggest that evaluating both near-term and long-term forecasts will be essential as research on forecasting methods accelerates. Second, while single approaches may perform reasonably well at either short or long forecast horizons, skillful predictions at intermediate time horizons may require a combination of information from spatial and temporal patterns. Intermediate time horizons pose challenges in other forecasting contexts as well. Weather forecasts based on regional-scale meteorological models are very effective for forecasting a week to ten days in advance, but then become largely uninformative. Forecasting these intermediate scales has been challenging in meteorology and will likely be challenging in ecology as well. While the recent emphasis on near-term iterative forecasting \citep{dietze_iterative_2018} is the logical and tractable starting point, we also need to build understanding and capacity for forecasting ecological dynamics across all temporal scales of interest.

\subsubsection*{Acknowledgements}

We thank Heather Lynch and Juan Manuel Morales for comments that improved early drafts of the paper. PBA was supported by the Utah Agriculture Experiment Station. EPW was supported by the Gordon and Betty Moore Foundation’s Data-Driven Discovery Initiative through Grant GBMF4563.

>>>>

\title{\textbf{ Matching the forecast horizon with the relevant spatial and temporal processes and data sources }}

@@

\noindent \emph{Acknowledgements:} We thank Heather Lynch and Juan Manuel Morales for comments that improved early drafts of the paper. PBA was supported by the National Science Foundation (DEB-1927282) and the Utah Agriculture Experiment Station (grant 1322). EPW was supported by the Gordon and Betty Moore Foundation’s Data-Driven Discovery Initiative (GBMF4563).

@@

@@

Most phenomenological, statistical models used to generate ecological forecasts take either a time-series approach, based on long-term data from one location, or a space-for-time approach, based on data describing spatial patterns across environmental gradients. Here we consider how the forecast horizon determines whether more accurate predictions come from the time-series approach, the space-for-time approach, or a combination of the two. We use two simulated case studies to show that forecasts for short and long forecast horizons need to focus on different ecological processes, which are reflected in different kinds of data. In the short-term, dynamics reflect initial conditions and fast processes such as birth and death, and the time-series approach makes the best predictions. In the long-term, dynamics reflect the additional influence of slower processes such as evolutionary and ecological selection, colonization and extinction, which the space-for-time approach can effectively capture. At intermediate time-scales, a weighted average of the two approaches shows promise. However, making this weighted model operational will require new research to predict the rate at which slow processes begin to influence dynamics.

@@

@@

Statistical models used for ecological forecasting generally rely on either time-series approaches or space-for-time substitutions. The time-series approach involves fitting models to long-term datasets to describe the temporal dynamics of a system.  We then use those dynamic models to make predictions about what will happen in the future. This approach is often used to study population or vital rate fluctuations as a function of weather \citep{dalgleish_climate_2011}, or primary production as a function of annual precipitation \citep{lauenroth_long-term_1992}. When time-series models are fit with typically short ecological data sets, they capture ``fast processes'' operating on interannual time-scales, such as birth, death, individual growth, small-scale dispersal events, and short-term responses to environmental conditions (Fig. \ref{fig:fast-slow-concept}). Statistical models built using this approach normally cover a limited spatial extent (but see \citealt{Hefley2017,kleinhesselink_response_2018} ADD Chevalier), and ignore slower processes, such as evolutionary adaptation or turnover in community composition, that could influence dynamics at longer time scales \citep{clark_ecological_2001}.

@@

Whether historical dynamics, contemporary spatial patterns, or some combination of the two will serve as the best source of 
information for forecasting may depend on how far into the future we are attempting to forecast \citep{harris_forecasting_2018}. 
This potential dependency on the ``forecast horizon'' (\textit{sensu} \citealt{Hyndman2018})
reflects lags in the response of ecological conditions to environmental change, shifts in the importance of ecological processes with
time scale \citep{levin_1992,rosenzweig_1995}, and differences between time-series and spatial gradients in the range of 
environmental conditions represented in observed data (Fig. \ref{fig:fast-slow-concept}).
At short forecast horizons (days to years), dynamics will reflect physiological and demographic responses and interactions
among the organisms present at a site more than temporal turnover of genotypes or species, 
environmental conditions are likely to stay within the range of historical variation, 
and the current state of the system is likely to capture the influence of unmeasured processes. As a result, 
for near-term forecasts time-series approaches may capture the key dynamics and provide accurate predictions.

In contrast, at long forecast horizons (decades to centuries),  environmental conditions that have not been 
historically observed are likely to not only occur but to persist long enough to drive significant turnover of genotypes and species 
through colonization and extinction as well as changes in the flux of energy and nutrients.  
At these long forecast horizons, the state of the system at the time the forecast is issued may be 
little help in predicting the future state. For the century-scale forecasts often featured in biodiversity and 
species-distribution modeling, space-for-time approaches may effectively capture the response of ecosystems to major shifts 
in climate over long periods, producing better long-term forecasts than time-series approaches. 
Using different modeling approaches for different forecast horizons is common in other disciplines.
For example, meteorological models for short-term weather forecasts differ substantially in spatial
and temporal resolution and extent from the global circulation models used to predict long-term changes
in climate.

Why not simply use process-based models to avoid the difficulties posed by these phenomenological models?
If we could accurately characterize all of the processes governing a system, then a model based on that understanding should make accurate predictions at all forecast horizons. Process-based models should also be more robust for making predictions outside of historically observed conditions and even beyond the conditions observed across spatial gradients, which will be especially important for making predictions in a future with increasingly novel combinations of environment and species interactions \citep{Williams2007}. Unfortunately, in most cases this approach is not currently feasible because we lack a detailed knowledge of all the complex and interacting processes influencing the dynamics of real ecological systems. Even if the general form of the models was known, estimating the high number of parameters and quantifying how they vary across ecosystems typically requires more data than is currently available even for well studied systems. Furthermore, the high complexity and corresponding parameter uncertainty of such models can increase predictive errors; simpler  time-series models may actually perform better \citep{ward_complexity_2014a}, though spatial replication can reduce the cost of complexity (ADD Chevalier).
-very difficult to parameterize 
-complexity is costly
-spatiotemporal stats models more realistic
-but seldom used for forecasting
-for now, models that rely in some way on phenomenological, statistical relationships will play an important role

Here we use two simulated case studies to 1) demonstrate that the best model-building approaches for ecological forecasting 
depend on the time horizon of the forecast, and 2) explore how time-series and space-for-time approaches might be 
combined via weighted averaging to make better forecasts at intermediate time scales. The first case study
focuses on how interspecific interactions affect the population dynamics of a focal species, and the second focuses on
an eco-evolutionary scenario. Our analyses show that: 
\begin{enumerate}
	\item For short-term forecasts, phenomenological time-series approaches are hard to beat, whereas longer-term forecasts require accounting for the influence of slow processes such as evolutionary and ecological selection as well as dispersal.
	\item Different kinds of data reflect the operation of different processes: longitudinal data capture autocorrelation and fast responses of current assemblages to interannual environmental variation, while data spanning spatial gradients capture the long-term outcome of interactions between fast and slow processes.  Whether predictive models should be trained using longitudinal or spatial data sets, or both, depends on the time-scale of the desired forecast.
	\item A key challenge for future research is determining the rate at which slow processes begin to influence dynamics.
\end{enumerate}

@@

In each case study, we simulated the effects of an increase in temperature on simple systems with known dynamics.
The truth is represented by a simulation model that is mechanistic for at least one important process, 
but we treat the data-generating model as unknown when analyzing the data and we assumed that perfectly recovering the mechanisms
it contains would not be possible in practice. We began each simulation
under a stationary distribution of annual temperatures, allowing the system to equilibrate; we call this the baseline phase. 
We then increased temperature progressively over a period of time, followed by a second period of stationary, now elevated, temperature. 
The objective was to forecast the response of the system to the temperature increase based on spatial and/or temporal data 
``sampled" from the simulation during the baseline period.

We made forecasts based on two phenomenological statistical models, each representing processes operating at different time scales. One statistical model represents the time-series or ``temporal approach.'' We correlated interannual variation in an ecological response with interannual variation in temperature at just one site. The other statistical model relies on a space-for-time substitution, which we call the ``spatial approach'' for brevity. We correlated the mean temperature with the mean of an ecological state or rate across many sites.  We compared forecasts from both statistical models to the simulated dynamics to determine how well the two approaches performed at different forecast horizons. We also assessed the potential for combining the information available in temporal and spatial patterns by using a weighted average of the forecasts from the temporal and spatial approaches optimized to best match the (simulated) observations. We then studied how the optimal model weights changed over time. We expected the temporal approach to best predict short-term dynamics, the spatial approach to best predict long-term dynamics, while the weighted model would show potential to provide the best forecasts at transitional, intermediate time scales. The three statistical models are described in Supporting Information (Appendix \ref{models}) and all code for both case studies is available for reviewers as Supporting Information (and will be archived at Zenodo upon acceptance).

@@

Conservation biologists and natural resource managers often need to anticipate the impact of environmental change on the abundance of 
endangered species, biological invaders, and harvested species. Although the managers may be primarily interested in just one focal species, skillful prediction might require considering interactions with many other species, greatly complicating the problem. But at what forecast horizon do altered species interactions become impossible to ignore? We explored this question using a metacommunity simulation model developed by \cite{alexander_lags_2018} to study how community responses to increasing temperature depend on the interplay between within-site demography and competitive interactions and the movement of species across sites. The model features Lotka-Volterra competitive interactions among plants within sites that are arrayed along an elevation and temperature gradient. Composition varies along the gradient because of a trade-off between growth rate and cold tolerance: cold sites are dominated by slow-growing species that can tolerate low temperatures, while warm sites are dominated by  fast-growing species that are cold intolerant. Multiple species can coexist within sites because all species experience stronger competition from conspecifics than from heterospecifics. Sites are linked by dispersal: a specified fraction of each species' offspring leaves the site where they were produced and reaches all other sites with equal probability. We provide a more detailed description of the simulation model in SI Appendix \ref{metacomm}.

@@

During the baseline period there were strong spatial patterns across the mean temperature gradient. 
Individual species, including our focal species, showed classic, unimodal ``Whittaker" patterns of abundances across the gradient (Fig. \ref{fig:species-patterns-models}A).
These spatial patterns are the basis for our spatial statistical model of the temperature-biomass relationship for our focal species (Fig. \ref{fig:species-patterns-models}A). 
In contrast to the strong spatial patterns, population and community responses to interannual variation in temperature within sites were weak.
At our focal site in the center of the gradient, the biomass of the focal species was quite insensitive to interannnual variation in temperature, but showed strong temporal autocorrelation (Fig. \ref{fig:species-patterns-models}B). Our temporal statistical model estimates this weak, linear temperature effect, along with the strong lag effect of biomass in the previous year.

We used both the temporal and spatial statistical models to forecast the effect of a temperature increase (Fig. \ref{fig:community-forecast-species}A) on the focal species' biomass at one location in the center of the temperature gradient. 
The predictions from these two models contrasted markedly, with the temporal 
statistical model predicting a large increase in biomass and the spatial statistical model predicting a decrease. Initially, the simulated abundances followed the increase predicted by the 
temporal model, but as faster-growing species colonized and increased in abundance at the focal site, the biomass of the focal species decreased, eventually falling below its baseline level (Fig. \ref{fig:community-forecast-species}B).

To combine information from the temporal and spatial statistical models into a single prediction, we fit a weighting parameter, $\omega$, which varies over time and is bounded between 0 and 1. At any time point, $t$, this weighted forecast is $\omega \cdot T(N_{t-1},K_t) + (1-\omega) \cdot S(K_t) $ where $T$ is the temporal statistical model, which depends on population size, $N$, and expected temperature, $K$, and $S$ is the spatial statistical model, which depends only on $K$ (see SI Appendix \ref{models} for a full description of the approach). The weighted model accurately predicts the simulated dynamics across the full forecast horizon (Fig. \ref{fig:community-forecast-species}B). It also shows that the most rapid shifts in the model weights occurred during the period when warm-adapted, faster growing species were increasing most rapidly in abundance (Fig. \ref{fig:community-forecast-species}C). However, the reason the weighted models works so well is that the weights were determined by fitting directly to the data. Unlike the forecasts from the spatial and temporal statistical models, we did not generate out-of-sample predictions from the weighted model; it merely provides a convenient way to quantify how rapidly dynamics shift from being dominated by interannual variation captured in the temporal model (time $t=0$ to $t \approx 1250$ in Fig. \ref{fig:community-forecast-species}B) to being dominated by the steady-state equilibrium captured by the spatial model (time $t \geq 2500$). A true forecast from the weighted model would require a method to determine the model weights \emph{a priori}.

The compositional turnover affecting our focal species also influences total biomass, linking community and ecosystem dynamics. We repeated our focal species analysis for total community biomass, and the results were similar: the temporal statistical model initially made the best forecasts immediately following the onset of the temperature increase, but as the identity and abundances of species at the study site changed, the model weights rapidly shifted to the spatial statistical model (SI Figs. \ref{fig:community-models-total} and \ref{fig:community-forecasts-total}).

@@

Evolutionary adaptation is a key uncertainty in predicting how environmental change will impact a focal population at a given location \citep{Hoffmann2011}. Like the shifts in species composition illustrated in the previous example, shifts in genotype frequencies can also influence dynamics and forecasts at different time scales. Although shifts in genotype frequencies at the population level are analogous to changes in species composition at the community level, the mechanisms are distinct: heterozygosity and genetic recombination have no analogue at the community level. We demonstrate how these processes influence short and long-term forecasts with a standard eco-evolutionary simulation model for a hypothetical annual plant population in which fecundity is temperature dependent, and different genotypes have different temperature optima (Fig. \ref{fig:evo_evo_spatial_temporal_models}A). Our model describes how the local density of each genotype changes between years, which depends on temperature and genotype densities in the previous year. Transient temporal dynamics are computed directly from the model; these dynamics are the basis for the temporal statistical model. To create a spatial gradient, we simulated the equilibrium density of each genotype in a series of local populations experiencing different mean temperatures. The pattern of equilibrium densities across the mean annual temperature gradient is the basis for our spatial statistical model: cold sites will be dominated by the cold-adapted homozygous genotype, warm sites will be dominated by the heat-adapted homozygous genotype, and intermediate sites will be dominated by the heterozygous genotype (Fig. \ref{fig:evo_evo_spatial_temporal_models}B). The full description of the eco-evolutionary simulation model is provided in SI Appendix \ref{eco-evo}.

@@

To demonstrate these dynamics, we simulated a diploid annual plant population at a colder than average site. During the baseline period, the population is dominated by the cold-adapted genotype. We used the simulated data from this baseline period to fit a temporal statistical model (Appendix \ref{models}) that predicts population growth rate as a function of annual temperature and population size (Fig. \ref{fig:evo_evo_spatial_temporal_models}C), assuming no knowledge of the underlying eco-evolutionary dynamics. We then imposed a period of warming, followed by a final period of higher stationary temperature (Fig. \ref{fig:forecast} top).

With the onset of warming, the population crashed as the cold-adapted genotype decreased in abundance. Eventually, frequencies of the heterozygous genotype and the warm-adapted homozygous genotype began to increase and the population recovered (Fig. \ref{fig:forecast} bottom). The temporal statistical model (solid blue line in Fig. \ref{fig:forecast}) accurately predicted the impact of the initial warming trend, but eventually became too pessimistic, while the spatial statistical model (solid red line in Fig. \ref{fig:forecast}) did not handle the initial trend but accurately predicted the eventual, new steady state (Fig. \ref{fig:forecast} bottom).

As in the community turnover example, we also fit a weighted average of predictions from the spatial and temporal statistical models, with the weights changing over time. This weighted model initially reflected the temporal model (decrease from $t=500$ to $t=600$), but then rapidly transitioned to reflect the spatial model ($t \geq 700$). The rapid transition in the weighting term, $\omega$, occurred during the period of most rapid change in genotype frequencies (Fig. \ref{fig:forecast_supp}). The weighted model's predictions look impressively accurate, but, as in the community turnover example, that is because we used the full, simulated time series to fit the weighting term. A true forecast would require an independent method to predict how the model weights shift over time.

@@

Ecological forecasts are typically made using either a space-for-time substitution approach based on models fit to spatial data or using dynamic models fit to time-series data. Our results demonstrate that these two approaches can make very different predictions about the future state of ecological systems. Which approach provides the most accurate forecasts depends on the forecast-horizon. In our simulations, time-series approaches performed best for short forecast horizons, whereas models based on spatial data made more accurate forecasts at long horizons. In addition, our simulations demonstrate extended transitional periods during which neither the time-series or the spatial approach was effective on its own. The challenge is determining what is ``short-term,'' what is ``long-term,'' and how to handle the many forecasts we need in ecology which fall in between. We have proposed that a weighted combination of the time-series and space-for-time approaches may produce better forecasts at these intermediate forecast horizons.

We designed our simulation studies to illustrate how the change in statistical model performance with increasing forecast horizon reflects differences in the types and scales of processes captured by spatial and temporal data sets. How could these hypotheses be tested with empirical data? The hypothesis that time-series models will be most effective for near-term forecasts already has empirical support, in the form of recent analyses of biodiversity forecasts at time scales from one to ten years \citep{harris_forecasting_2018}. The result should not be surprising, since local time-series data capture demographic processes, lagged effects, and responses of current assemblages to small changes in environmental conditions. In addition, the state of the system in the near future depends heavily on the current state. Since short-term forecasts do not typically require extrapolating into novel conditions, a model based on the historical range of variation which incorporates lags and accurate initial conditions is likely to be successful.

@@

The greatest empirical challenge will be testing our hypothesis that a weighted average of spatial and temporal statistical models will make the best forecasts at intermediate time scales. There are two problems: finding appropriate data and determining the model weights \emph{a priori}. Many data sets have both a longitudinal and spatial dimension, but we could not think of one which also featured a clear ecological response to significant environmental change. Surely such datasets exist, and we hope researchers who work with them will test our proposed weighted model. Determining model weights may be more difficult. In our simulations, we fit the weights directly to the simulated data, which is impossible to do for actual forecasting when the future is unknown. We need new theory or empirical case studies in order to assign these weights \emph{a priori}.

@@

An alternative to a weighted combination of predictions from time-series and space-for-time models is to rely on fully process-based models. 
% mention difficulty of dealing with error propogation?
As a result, models used for ecological forecasting will include at least some phenomenological components. But that does not mean that phenomenological forecast models do not benefit from process-based understanding. The message from our simulations is that different processes should be considered for different forecast time-scales, and this can be done by fitting models to different kinds of datasets. Even when process-level understanding does not enable a fully mechanistic model, it can improve the specification of phenomenological models.

@@

Our results have important implications for the emerging field of ecological forecasting. First, they suggest that evaluating model performance at both short and long forecsat horizons will be essential as research on forecasting methods accelerates. Second, while single approaches may perform reasonably well for either short or long horizons, skillful predictions at intermediate forecast horizons may require a combination of information from spatial and temporal statistical models. Intermediate time horizons pose challenges in other forecasting contexts as well. Weather forecasts based on regional-scale meteorological models are very effective for forecasting a week to ten days in advance, but then become largely uninformative. Forecasting these intermediate scales has been challenging in meteorology and will likely be challenging in ecology as well. While the recent emphasis on near-term iterative forecasting \citep{dietze_iterative_2018} is the logical and tractable starting point, we also need to build understanding and capacity for forecasting ecological dynamics across all forecast horizons of interest.

\subsubsection*{Data accessibility statement:} The manuscript contains no original data. All computer code is uploaded for review in a zip file. Upon acceptance of the manuscript, code will be archived at Zenodo.
